Detecting Stylistic Variation in Pop Production using the Spotify API

### 2023-11-12 outputs/paper/paper.qmd
Paper tone/style edits, fixed the regression

We find that danceability is the strongest predictor of whether or not a song has been produced by Antonoff. Furthermore, when the variance is considered along with other features of the dataset, the variance tends to be strongest classifier of whether a song was produced by Antonoff. Danceability in concert with other features in the dataset provides the best predictions.

Against the odds, we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools. The remainder of this paper is structured as follows: @ref-background...

# Background {#sec-background}

Shortly after the release of Taylor Swift's *Midnights* album, a video of the Youtuber Caleb Gammen, in which he guesses which songs on Midnights were produced by Antonoff within seconds [@citeTweet], went viral. The tweet reads: "im able to instantly detect if jack antonoff worked on a song due to a visceral hatred of his production style". He nails it. Only one song on the album wasn't produced by Antonoff, and he guesses it.

Antonoff began his career as a guitarist in fun., and is the frontman in Bleachers. He has produced and co-produced music for Lorde, Taylor Swift, Florence and the Machine, The Chicks, Clairo, The 1975, Grimes, Zayn, Pink, Lana Del Rey, Olivia Rodrigo, the Minions: The Rise of Gru Soundtrack, and more. Antonoff is prolific, to the extent that it is starting to bother people. As Andrew Marantz writes in the New Yorker: "When his band releases an album, the world responds politely. When he produces one by Lorde or Lana Del Rey or Taylor Swift, the world wobbles on its axis" [@citeMarantz].Can Spotify's audio features dataset help us determine what makes Antonoff music so Antonoff-y?

I spoke to Gamman to better undertstand makes a song Antonoff-ish, and to try to understand his reaction to it.

I also spoke with Glen McDonald, an engineer at Spotify, from whom I learned that the `valence` and `danceability` variables in the Spotify API were created by hiring college interns and having them tag whether songs were positive or danceable.

However, logistic regression assumes a linear relationship between the dependent and predictor variables, and is sensitive to multi-collinearity. In music data, we would expect for certain variables to be correlated (for instance, loudness and energy). We also cannot assume that an artists' style can be measured in the linear growth of these metrics. Although the logistic regression has a strong accuracy, it may be detecting a change in genre, rather than a change in style, across artists whose sound turns more and more pop while collaborating with Antonoff.

In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness.

@@

The audio feature data was pulled using Spotify API, which I accessed using the `spotifyr` package in R [@citeSpotify]. I used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski were included, since they've never collaborated with Jack Antonoff but are more or less part of the same pop sphere.

I also scraped the discographies of comparable contemporary pop producers' Wikipedias. The producers include Epworth, Rechstaid, Max Martin, Nowels, Joel Little, and Greg Kustin. (cite Carl here?) These producers were chosen based on their popularity, and also because some of them collaborated with artists who also collaborate with Antonoff.

@@

Of the artists who did collaborate with Antonoff, I chose these artists because they've produced multiple albums, and at least one album with Antonoff, as opposed to just individual songs.

We created a binary variable which contains 1 if Jack Antonoff produced or co-produced the song, and a 0 if it was produced by someone else. The information underpinning this discography was drawn from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

@@

It's hard to gauge exactly what the variables `danceability` or `valence` mean, even using the documentation. For instance, `valence` is described as "A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)". `Danceability` is described as "how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable". Neither of these definitions fully cover what metrics and thresholds go into something like `danceability`.

In total 3738 tracks from 828 different artists. We focus on identifying the styles of 7 producers: Jack Antonoff, Max Martin, Joel Little, Ariel Rechtshaid, Greg Kurstin, Paul Epworth and Rick Nowels. If tracks were produced by producers other than the main producers included in our study, they were marked as "other". We removed tracks from live performances, karaoke editions, international versions or remix albums. Where a deluxe album was available, we removed tracks belonging to the origin to avoid repetition.

@@

In @fig-quadvalence, it looks like songs produced by Antonoff are a little more clustered towards the lower end of valence, and across the spectrum of energy, although the variance is large.

```{r}
#| label: fig-quaddanceability2
#| fig-cap: Danceability and Energy Quadrant
#| echo: false
#| warning: false

ggplot(av, aes(x = energy, y = danceability)) +
   geom_point(aes(color = as.factor(is_antonoff)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$danceability),
              color = "gray") +
  theme_minimal()
```

In @fig-quaddanceability2, it looks like Antonoff-produced songs err towards higher danceability, across the spectrum of energy.

These quadrants are a start, but they don't suggest any defining Antonoff-iness characteristic underlying each of those blue points, which are scattered pretty broadly.

By comparing cumulative distribution functions of each variable based on whether a song was produced by Antonoff, we can see that some of these variables do capture some sort of underlying Antonoff-iness (@fig-cdf). For instance, while liveness, instrumentalness, and speechiness are relatively similar, we can see clear differences in danceability, energy, loudness, acousticness, and valence.

@@

Another way to gauge the Antonoff-ification of pop music could be to look at how artists' sounds have changed over time, and whether there is any correlation with Antonoff. These plots are here to get a glimpse at pop's evolution over time, and where Antonoff fits in that picture. Each point represents a song.

@@

In @fig-energy, it looks like there's a downward trend in energy scores in artists who work with Antonoff, but that these downward trends are also somewhat in line with the artists' energy trajectories anyway, which could be indicative of broader trends in pop. Specifically, Lana Del Rey and Taylor Swift. They have enough albums recorded prior to collaborations with Antonoff to be able to observe this. Sharon Von Etten, Maggie Rogers, and Mitski, who have never worked with Antonoff, all have upward trends.

@@

There aren't any significant correlations in danceability (@fig-danceability) .

@@

In @fig-valence, it looks like Lorde has the only significant change in valence scores, which tracks with my understanding her albums. *Melodrama* had piano and break ups, and *Solar Power* had a music video of Lorde prancing on the beach in a yellow bikini with Phoebe Bridgers and Clairo.

@@

Again in @fig-acousticness, Lorde's acoustic Solar Power album jumps out. Taylor Swift, Lorde and HAIM also have significant-seeming upward trends in `acousticness`, while Mitski, Sharon Von Etten and Maggie Rogers see decreasing `acousticness`.

@@

In @fig-instrumentalness, St. Vincent is the only plot with any significant trend in `instrumentalness`.

@@

Finally, in @fig-tempo, there aren't many significant trends in tempo, except for Antonoff's collaborations with Lana Del Rey and Taylor Swift.

We've observed subtle distinctions in variables associated with songs produced by Jack Antonoff, as evidenced by our cumulative distribution functions, quadrant analyses, and minor trends in the evolution of artists' sound characteristics, particularly energy. Employing a logistic regression will then: a) determine the likelihood of a song being produced by Antonoff based on these identified variables, and b) discern which variables are most influential in shaping this classification.

@@

@@

We used the `tidymodels` to fit the model [@citetidymodels]. To deal with the imbalanced classes, we downsampled the majority class using the **`step_downsample()`** function from the **`themis`** package. Then we used 5-fold cross-validation with stratification by the **`jack`** variable to estimate the performance of the model.

The `workflow_set()` function lets us hold different sets of feature engineering together and use tuning to evaluate them all at once. We need to evaluate two elements: a basic model, and the model that was downsampled using the `themis` package.

We used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity. After testing both basic and downsampled versions of the model, the basic version yielded much higher accuracy score (0.918 basic versus 0.743 downsampled), although a lower roc auc score (0.746 basic versus 0.821 downsampled).

The version with the original data remains consistently high in accuracy, even with increased regularization. It starts low with in mean log loss and sensitivity, but increases with increased regularization. Compared to the downsampled data, which starts high but drops significantly in accuracy and sensitivity, we chose to use the original data.

@@

The model's accuracy score is 0.918, and the ROC AUC score is 0.746.

|                        | Actually Positive (1) | Actually Negative (0) |
|------------------------|-----------------------|-----------------------|
| Predicted Positive (1) | 848                   | 76                    |
| Predicted Negative (0) | 1                     | 0                     |

@@

@tbl-importance shows the different variables' importance on determining whether a song was had any Antonoff involvement.

@@

`Danceability`, with a positive importance score of 4.142, had the highest level of importance in determining whether a song was produced or co-produced by Jack Antonoff. `Valence` comes in second at 2.559, with negative score. The third, `instrumentalness` correlates negatively with Antonoff tracks with a a negative score of 2.309.

@@

@tbl-mostdanceable shows us the 20 songs in Jack Antonoff's discography with the highest `danceability` scores. Bleachers takes up 35% of this list, and Taylor Swift takes up 40% of this list. St Vincent, The Chicks, Lorde, and Florence and the Machine also haves songs up there.

@@

Our results indicate that Antonoff's music, characterized by its danceability score, exhibits a significant positive correlation with factors such as speechiness, loudness, and energy, while showing a negative correlation with acousticness, and a less pronounced positive correlation with valence, compared to other music producers. This profile---marked by speechiness, loudness, energy, synthetic textures, and a blend of positivity with a touch of nostalgia---aligns with both detailed auditory analysis and critical assessments of Antonoff's sound.

@@

Our analysis reveals that the amalgamation of crowdsourced metrics and machine-listening variables offers a nuanced approximation of musical style. Interestingly, the subjective nature of crowdsourced metrics, which might appear rudimentary, plays a crucial role in shaping a more intuitive and embodied understanding of music. This synergy between human subjectivity and algorithmic precision not only mirrors the multifaceted nature of musical perception, but also enriches our metric's capacity to capture the distinctive essence of an artist's style.

>>>>

We find that `danceability` is the strongest predictor of whether or not a song has been produced by Antonoff. Furthermore, when `danceability`'s correlation scores with other features in the dataset are considered, the variance in correlation scores tends to be strongest classifier of whether a song was produced by Antonoff. Danceability in concert with other features in the dataset provides the best predictions.

Against the odds, we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools. The remainder of this paper is structured as follows:

@ref-data

@ref-Model

@ref-Results

@ref-Discussion

@ref-References

@@

We collected data from the Spotify API, which we accessed using the `spotifyr` package in R [@citeSpotify]. We used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski were included to represent other contemporary pop artists with overlapping fan bases who have never collaborated with Antonoff. Of the artists who did collaborate with Antonoff, we chose artist who have produced multiple albums, and at least one album with Antonoff.

We acquired the discographies of comparable contemporary pop producers' by scraping their respective discography Wikipedia pages. The producers include Epworth, Rechstaid, Max Martin, Nowels, Joel Little, and Greg Kustin. These producers were chosen based both on their popularity, and because some have collaborated with the same artists as Antonoff.

@@

We created a binary variable which contains 1 if Jack Antonoff produced or co-produced the song, and 0 if it was produced by somebody else. The information underpinning this discography was drawn from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

In total, we collected 3738 tracks from 828 different artists. If tracks were produced by producers other than the main producers included in our study, they were marked as "other".

```{r}
#| label: tbl-producercounts
#| tbl-cap: Counts of Tracks Per Producer in Original Dataset
#| echo: false
#| warning: false

library(kableExtra)

tbl <- table(lr_data$producer)

tbl |>
  kable(col.names = c("Producer", "Number of Tracks"))

@@

@@

By analyzing the cumulative distribution functions of various audio features, particularly when categorizing songs based on whether they were produced by Jack Antonoff (@fig-cdf), distinct patterns emerge. Notably, while attributes such as liveness, instrumentalness, and speechiness exhibit relative homogeneity, there are slight disparities in features like danceability, energy, loudness, acousticness, and valence.

@@

Examining the temporal evolution of artists' sounds, especially in correlation with their collaboration with Antonoff, provides insights into the broader trends in pop music. Each point in the following plots represent one track.

@@

In @fig-energy, a downward trend in the energy scores is observed among artists collaborating with Antonoff. However, this trend appears consistent with the overall trajectories of these artists' works, suggesting broader trends in the pop genre, as exemplified by Lana Del Rey and Taylor Swift. Sharon Von Etten, Maggie Rogers, and Mitski, display upward energy trajectories in their musical careers.

@@

No significant trends are evident in danceability across the artists' collaborations with Antonoff, as demonstrated in (@fig-danceability) .

@@

@fig-valence reveals a notable shift in valence scores in Lorde's discography, where her tracks display a slight upward trend and a greater range in valence scores over time.

@@

@fig-acousticness underscores a trend toward increased acousticness in the works of Taylor Swift, Lorde, and HAIM, contrasted with declining trends in artists like Mitski, Sharon Von Etten, and Maggie Rogers.

@@

@fig-instrumentalness exhibits no significant correlations.

@@

Finally, tempo trends, as shown in Figure @fig-tempo, are generally stable across artists, with notable exceptions in collaborations involving Lana Del Rey and Taylor Swift.

We've observed subtle distinctions in variables associated with songs produced by Jack Antonoff, as evidenced by our cumulative distribution functions and minor trends in the evolution of artists' sound characteristics, particularly energy. This study sets the stage fora) employing logistic regression to quantify the likelihood of Antonoff's involvement in a song based on these audio features, b) classifying tracks produced by a wider range of producers using a Random Forest, c) identify the most influential variables in this classification.

@@

The aim of the logistic regression is to determine whether we can predict Antonoff's influence in artists' sound. We select a total of 1465 tracks from the dataset from 12 artists: Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, Mitski, and Bleachers (Antonoff's own band)..

@@

We used the `tidymodels` to fit the model [@citetidymodels]. To deal with the imbalanced classes, we downsampled the majority class using the **`step_downsample()`** function from the **`themis`** package. Then we used 5-fold cross-validation with stratification by the **`is_antonoff`** variable to estimate the performance of the model.

@@

The model's accuracy score is 0.757, and the ROC AUC score is 0.819. The precision score is 0.3851, meaning that approximately 38.51% of songs predicted to be produced by Antonoff were actually produced by him. The recall of 0.7561 indicates that about 75.61% of songs produced by Antonoff were actually correctly identified in the model.

|                        | Actually Positive (1) | Actually Negative (0) |
|------------------------|-----------------------|-----------------------|
| Predicted Positive (1) | 62                    | 99                    |
| Predicted Negative (0) | 20                    | 308                   |

@@

@tbl-importance shows variables' respective importance in determining whether a track was produced by Antonoff.

@@

`Danceability`, with a positive importance score of 4.142, had the highest level of importance in determining whether a song was produced or co-produced by Jack Antonoff. `Valence` was the second most important, with a negative score of 2.559. The third, `instrumentalness` correlates negatively with Antonoff tracks with a negative score of 2.309.

@@

In the 20 songs in Jack Antonoff's discography with the highest `danceability` scores, Bleachers tracks account for 35% of this list, and Taylor Swift tracks account 40% of this list (@tbl-mostdanceable). St Vincent, The Chicks, Lorde, and Florence and the Machine are also featured.

@@

Our logistic regression model was not able to strongly predict whether or not a song was produced by Jack Antonoff using the Spotify API's audio features data. These data are limited, providing static metrics that stand in for entire tracks, which may not be detailed enough for intra-genre classification. However, when classifying the tracks, `danceability` was the strongest predictor.

Our Random Forest results, which were moderately strong at classifying tracks, indicate that Antonoff's music, characterized by its `danceability` score, exhibits a significant positive correlation with factors such as `speechiness`, `loudness`, and `energy`, while showing a negative correlation with `acousticness`, and a less pronounced positive correlation with `valence`, compared to other music producers. This profile---marked by speechiness, loudness, energy, synthetic textures, and a blend of positivity with a touch of nostalgia---aligns with both detailed auditory analysis and critical assessments of Antonoff's sound.

@@

Our analysis reveals that the amalgamation of crowdsourced metrics (danceability) and machine-listening variables offers a nuanced approximation of musical style. Interestingly, the subjective nature of crowdsourced metrics, which might appear rudimentary, plays a crucial role in shaping a more intuitive and embodied understanding of music. This synergy between human subjectivity and algorithmic precision not only mirrors the multifaceted nature of musical perception, but also enriches our model's capacity to capture the essence of an artist's style.
